\begin{preface}
It does not matter what you order, the mainstay of any meal will be salt and pepper. Salt and pepper are so phenomenally successful in the herbal and spice arena. They arrogantly sit on the table, knowing that what ever you want to prepare you are going to require one of us. "Me myself, and my dear friend Pep. Isn't that right?" asked Salt. "Yes you are salt, have you been busy this morning?" replied Pep. "Oh I was ALL over that!" salt replied. But there are other herbs and spices: oregano, thyme, rosemary, cumin, paprika, \dots they are all lined-up in the cupboard waiting. Sometime the door cupboard door opens, they see salt and pepper on the table. "Look at them, the bastards! What have they got! What have they got that we don't have! I just don't get it!", said Oregano. "I don't get it either.", replied Thyme. "Neither do I.", followed Rosemary quickly. "I can't even see them from here, what do they look like?", asked Cumin. "You've been out of the cupboard Cumin?", asked Oregano. "Once, i think, for bolognese. Euhm they said it was an experiment, that was in 1992, and i don't think it worked. But i remain hopeful." answered Cumin, desperately smiling. "What about you Sage? Have you been out of the cupboard?" asked Oregano. "I actually used to live with them in their old house, and euhm then they moved me here, and eum put me back in this cupboard. No" he replied. "What about you paprika?", "I felt out, once, they just put me back the wrong way around. 7 years i was turned around. Every day i pray, i pray for it to be goelash day. But it is never goalash day! I still got the plastic on my head why the fuck did they buy me, i don't understand it!" with a confused voice. "What about you? Who are you on the end?" asked Oregano. "My name is five spice!" Five Spice responded confidently. Have you been out of the cupboard wild spice? "I'm no one spice, I have five spice. I'm five times as good as all of you!" Five Spice replied proudly. "Have you ever been out of the cupboard?" asked Oregano. "I have been used in recipes for thousands of years!" "But have you ever been out of the cupboard?!" asked Oregano again, slightly irritated. "No..." said Five Spice disappointed. "Excuse me. I'm sorry to interrupt all of this nonsense, I'm so tired listening to all of you whining because the fact is i have been in this cupboard longer than all of you!; put together." "Sorry, who are you?" said Oregano in a surprised, yet annoyed, way. "John West, tuna, nice to meet you." he replied.\\

\hfill - \textsf{Michael Mcintyre}
\end{preface}
